Cryptanalysis of \MORUS

### 2018-05-10 morusAC_01_Introduction.tex
updated contributions and discussion to reflect the fact that the correlation is between plaintext and ciphertext bits, rather than ciphertext bits with a fixed plaintext

\subsubsection*{Related Work.}
In the \MORUS submission document, the designers discuss the security of \MORUS against
several attacks, including algebraic, differential, and guess-and-determine attacks.
The main focus is on differential properties, and not many details are given for other attack vectors.
In third-party analysis,
Mileva et~al.~\cite{balkancryptsecMilevaDV15} propose a distinguisher in the nonce-reuse setting and practically evaluate the differential behaviour of toy variants of \MORUS.
Dwivedi et~al.~\cite{cryptoeprint:2016:1053} discuss the applicability of SAT solvers for state recovery, but the resulting complexity of $2^{370}$ for \MORUS[640] is well beyond the security claim.
Dwivedi et~al.~\cite{secryptDwivediMW17} also propose key-recovery attacks for \MORUS[1280] if initialization is reduced to 3.6 out of 16 steps, and discuss the security of \MORUS against internal differentials and rotational cryptanalysis.
Salam et~al.~\cite{trustcomSalamSBDPW17} apply cube attacks to obtain distinguishers for up to 5 out of 16 steps of the initialization of \MORUS[1280] with negligible complexity.
Additionally, 
Kales et~al.~\cite{cryptoeprint:2017:1137} and
Vaudenay and Viz\'{a}r \cite{cryptoeprint:2017:1147}
independently propose state-recovery and forgery attacks on \MORUS in a nonce-misuse setting with negligible data and time complexities.
Finally, a keystream correlation similar in nature to our main attack was uncovered by Minaud \cite{sacryptMinaud14} on the authenticated cipher \cipher{AEGIS} \cite{AEGIS,sacryptWuP13}, another CAESAR finalist, which shares the same overall structure as \MORUS, but uses a very different, \cipher{AES}-based state update function.

@@

In more detail, our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking ciphertext bits across five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
As a consequence, when encrypting a fixed plaintext segment multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the correlation depends purely on the plaintext, and not on the secret key of the cipher.
In principle, this property could be used to recover unknown bits of a plaintext encrypted a large number of times, provided an initial segment of the plaintext is known.
For \MORUS[1280] with 256-bit keys, the linear correlation is $2^{-76}$
and can be exploited using about $2^{152}$ encrypted blocks.

To the best of our knowledge, this is the first attack on full \MORUS in the nonce-respecting setting.
We also note that rekeying does not prevent the attack: the biases are independent of the secret encryption key and nonce, and can be exploited as long as a given plaintext segment is encrypted sufficiently often, regardless of whether each encryption uses a different key.

@@

\subsubsection*{Outline.}
This paper is organized as follows.
We first provide a brief description of \MORUS in \Cref{sec/Preliminaries}.
In \Cref{sec/introminimorus}, we introduce \MiniMORUS, an abstraction of \MORUS based on a certain class of rotational symmetries.
We analyze this simplified scheme in \Cref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \Cref{sec/fulltrails}, showing a correlation in the keystream over 5 steps,
and discuss the implications of our observation for the security of \MORUS in \Cref{sec/Discussion}.
In \Cref{sec/IniFin}, we present our results on the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
Finally, we conclude in \Cref{sec/Conclusion}.

>>>>

\subsubsection*{Related Work.}
In the \MORUS submission document, the designers discuss the security of \MORUS against
several attacks, including algebraic, differential, and guess-and-determine attacks.
The main focus is on differential properties, and not many details are given for other attack vectors.
In third-party analysis,
Mileva et~al.~\cite{balkancryptsecMilevaDV15} propose a distinguisher in the nonce-reuse setting and practically evaluate the differential behaviour of toy variants of \MORUS.
Shi et~al.~\cite{aiieShiGLZ16} analyze the differential properties of the finalization reduced to 2 out of 10 steps, but find no attacks.
Dwivedi et~al.~\cite{cryptoeprint:2016:1053} discuss the applicability of SAT solvers for state recovery, but the resulting complexity of $2^{370}$ for \MORUS[640] is well beyond the security claim.
Dwivedi et~al.~\cite{secryptDwivediMW17} also propose key-recovery attacks for \MORUS[1280] if initialization is reduced to 3.6 out of 16 steps, and discuss the security of \MORUS against internal differentials and rotational cryptanalysis.
Salam et~al.~\cite{trustcomSalamSBDPW17} apply cube attacks to obtain distinguishers for up to 5 out of 16 steps of the initialization of \MORUS[1280] with negligible complexity.
Additionally, 
Kales et~al.~\cite{cryptoeprint:2017:1137} and
Vaudenay and Viz\'{a}r \cite{cryptoeprint:2017:1147}
independently propose state-recovery and forgery attacks on \MORUS in a nonce-misuse setting with negligible data and time complexities.
Finally, a keystream correlation similar in nature to our main attack was uncovered by Minaud \cite{sacryptMinaud14} on the authenticated cipher \cipher{AEGIS} \cite{AEGIS,sacryptWuP13}, another CAESAR finalist, which shares the same overall structure as \MORUS, but uses a very different, \cipher{AES}-based state update function.

@@

In more detail, our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking plaintext and ciphertext bits spanning five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
%As a consequence, when encrypting a fixed plaintext segment multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the correlation does not depend on the secret key of the cipher.
In principle, this property could be used as a known-plaintext distinguisher, or to recover unknown bits of a plaintext encrypted a large number of times.
For \MORUS[1280] with 256-bit keys, the linear correlation is $2^{-76}$
and can be exploited using about $2^{152}$ encrypted blocks.

To the best of our knowledge, this is the first attack on full \MORUS in the nonce-respecting setting.
We note that rekeying does not prevent the attack: the biases are independent of the secret encryption key and nonce, and can be exploited for plaintext recovery as long as a given plaintext segment is encrypted sufficiently often, regardless of whether each encryption uses a different key.
A notable feature of the linear trail underpinning our attack is also that it does not depend on the values of rotation constants: a very similar trail would exist for most choices of round constants.

@@

\subsubsection*{Outline.}
This paper is organized as follows.
We first provide a brief description of \MORUS in \Cref{sec/Preliminaries}.
In \Cref{sec/introminimorus}, we introduce \MiniMORUS, an abstraction of \MORUS based on a certain class of rotational symmetries.
We analyze this simplified scheme in \Cref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \Cref{sec/fulltrails}, showing a correlation in the keystream over 5 steps,
and discuss the implications of our observation for the security of \MORUS in \Cref{sec/Discussion}.
In \Cref{sec/IniFin}, we present our results on the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
We conclude in \Cref{sec/Conclusion}.

### 2018-05-10 morusAC_03_MiniMorus.tex
Some more copyediting
updated contributions and discussion to reflect the fact that the correlation is between plaintext and ciphertext bits, rather than ciphertext bits with a fixed plaintext

To simplify the description of the attack, we assume all plaintext blocks are zero. However the same results would hold as long as the plaintext is fixed. Recall that the inner state of the cipher consists of five $4w$-bit registers $S_0,\dots,S_4$, each containing four $w$-bit words.

@@

To build our linear trail, we start in a linear combinations of bits within a single register.
\begin{definition}
Recall that $w$ denotes the word size in bits, and $4w$ is the size of a register. A linear combination of the form:
\[
S^t_{i,j(0)} \oplus S^t_{i,j(1)} \oplus \dots \oplus S^t_{i,j(k)}
\]
is said to be \emph{symmetric} iff the set of bits $S^t_{i,j(0)}, \dots, S^t_{i,j(k)}$ is left invariant by a circular shift by $w$ bits; that is, iff:
\[
\{j(i) : i\leq k\} = \{j(i) + w \text{\rm{} mod } 4w : i\leq k\}.
\]
\end{definition}
\emph{Example.} The following linear combination is symmetric for \MORUS[640], i.e. $w = 32$:
\begin{equation}
S^t_{0,0} \oplus S^t_{0,32} \oplus S^t_{0,64} \oplus S^t_{0,96}.
\label{eq:symmetric}
\end{equation}

>>>>

To simplify the description of the attack, we assume all plaintext blocks are zero. This assumption will be removed in \Cref{subsec:variable}, where we will show that plaintext bits only contribute linearly to the trail. Recall that the inner state of the cipher consists of five $4w$-bit registers $S_0,\dots,S_4$, each containing four $w$-bit words.

@@

To build our linear trail, we start with a linear combinations of bits within a single register.
\begin{definition}
Recall that $w$ denotes the word size in bits, and $4w$ is the size of a register. A linear combination of the form:
\[
S^t_{i,j(0)} \oplus S^t_{i,j(1)} \oplus \dots \oplus S^t_{i,j(k)}
\]
is said to be \emph{symmetric} iff the set of bits $S^t_{i,j(0)}, \dots, S^t_{i,j(k)}$ is left invariant by a circular shift by $w$ bits; that is, iff:
\[
\{j(i) : i\leq k\} = \{j(i) + w \text{\rm{} mod } 4w : i\leq k\}.
\]
\end{definition}
\emph{Example.} The following linear combination is symmetric for \MORUS[640], i.e. $w = 32$:
\begin{equation}
S^t_{0,0} \oplus S^t_{0,32} \oplus S^t_{0,64} \oplus S^t_{0,96}.
\label{eq:symmetric}
\end{equation}

### 2018-05-10 morusAC_05_Full_trails.tex
updated contributions and discussion to reflect the fact that the correlation is between plaintext and ciphertext bits, rather than ciphertext bits with a fixed plaintext

It may be observed that message blocks in the \StateUpdate{} function only contribute linearly to the state, and in that regard play a role similar to key bits in an SPN cipher; and indeed in SPN ciphers, key bits contribute linearly to linear trails \cite{eurocryptMatsui93}.

>>>>

It may be observed that message blocks in the \StateUpdate{} function only contribute linearly to the state, and in that regard play a role similar to key bits in an SPN cipher; and indeed in SPN ciphers, it is the case that key bits contribute linearly to linear trails \cite{eurocryptMatsui93}. In this light the previous result may not be surprising.

### 2018-05-10 morusAC_06_Discussion.tex
page break optimization
added subsection 5.3 about variable plaintext
updated contributions and discussion to reflect the fact that the correlation is between plaintext and ciphertext bits, rather than ciphertext bits with a fixed plaintext

@@

We emphasize that the correlation we uncover between ciphertext bits is
\emph{absolute}, in the sense that it does not depend on the encryption
key, or on the nonce.  This is the same situation as the keystream correlations
in AEGIS~\cite{sacryptMinaud14}.  As such, they can be leveraged to
mount an attack in the broacast setting, where the same message is
encrypted multiple times with different IVs and potentially different
keys~\cite{DBLP:conf/fse/MantinS01}.  In particular, the broadcast setting appears
in practice in man-in-the-browser attacks against HTTPS connections
following the BEAST model~\cite{duong2011here}.  In this scenario, an
attacker uses Javascript code running in the victim's browser (by
tricking the victim to visit a malicious website) to generate a large
number of request to a secure website.  Because of details of the HTTP
protocol, each request includes an authentication token to identify the
user, and the attacker can target this token as a repeated plaintext.
Concretely, correlations in the RC4 keystream have been exploited in this
setting, leading to the recovery of authentication cookies in
practice~\cite{DBLP:conf/uss/AlFardanBPPS13}.

@@

Moreover, the existence of this trail does hint at some weakness in the
design of \cipher{MORUS}. Indeed, a notable feature of the trail is that
the values of rotation constants are mostly irrelevant: a similar trail
would exist for most choices of the constants. That it is possible to
build a trail that ignores rotation constants may be surprising. This
would have been prevented by adding a word-wise rotation to one of the
state registers at the input of the ciphertext equation.

>>>>

%\pagebreak[4]

@@

We emphasize that the correlation we uncover between plaintext and ciphertext bits is
\emph{absolute}, in the sense that it does not depend on the encryption
key, or on the nonce.  This is the same situation as the keystream correlations
in AEGIS~\cite{sacryptMinaud14}.  As such, they can be leveraged to
mount an attack in the broacast setting, where the same message is
encrypted multiple times with different IVs and potentially different
keys~\cite{DBLP:conf/fse/MantinS01}.  In particular, the broadcast setting appears
in practice in man-in-the-browser attacks against HTTPS connections
following the BEAST model~\cite{duong2011here}.  In this scenario, an
attacker uses Javascript code running in the victim's browser (by
tricking the victim to visit a malicious website) to generate a large
number of request to a secure website.  Because of details of the HTTP
protocol, each request includes an authentication token to identify the
user, and the attacker can target this token as a repeated plaintext.
Concretely, correlations in the RC4 keystream have been exploited in this
setting, leading to the recovery of authentication cookies in
practice~\cite{DBLP:conf/uss/AlFardanBPPS13}.

@@

\subsubsection{Design considerations.}
The existence of this trail does hint at some weakness in the
design of \cipher{MORUS}. Indeed, a notable feature of the trail is that
the values of rotation constants are mostly irrelevant: a similar trail
would exist for most choices of the constants. That it is possible to
build a trail that ignores rotation constants may be surprising. This
would have been prevented by adding a word-wise rotation to one of the
state registers at the input of the ciphertext equation.
